**Machine learning identifies unique taxa differentiating proximal and distal

### 2017-06-07 submission/manuscript.Rmd
methods

### Bacterial OTUs associated with cancer are found in healthy individuals 
Some of the healthy patients have Fusobacterium OTUs. The most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subspecies animalis_. It is the only OTU to align with a nucleatum strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. ((cite joe and niels work somewhere in here, maybe in the discussion?)). The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample. ((Table or something?)). OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides. Samples that were positive for _P. asacharolytica_ ranged in abundances from 0.01% - 16%. Finally, another commonly cancer-associated species, _Parvimonas micra_ was identified to be most similar to OTU248. Only 4 individuals had this bacteria present in their colons, three in the left mucosa and one in the right mucosa. Thus, potentially oncogenic OTUs can be found in our sample set of 20 healthy individuals.

@@

We detected _F. nucleatum_ and _P. asacharolytica_ in 7 and 5 of our subjects, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans (cite Niel) and have oncogenic properties in cell culture and in mice (cite). Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the left mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics, it might not be surprising that it resides in the less-oygen-rich left bowel (cite something). In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on right-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum (cite Mima 2016). Of the (8) (40%) individuals positive for _F. nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumen and left mucosa. That said, in a previous study of 1,102 colorectal cancer cases, only 138 (11%) of them contained detectable _F. nucleatum_ at all (Mima 2016). In a separate study of patients with inflammatory bowel disease (IBD), 50% of 22 IBD patients and 17.6% of 34 people without disease contained _F. nucleatum_ in mucosal biopsies (Strauss 2011). These studies have the added caveat of being sampled from prepared or cleaned colons. Thus, the presence of _F. nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers. Because of the spatial distribution of the _F. nucleatum_ in our sample set, we cannot develop a model for the role of _F. nucleatum_ in the healthy colon. Indeed, rather than the _F. nucleatum_ initiating tumorigenesis, these bacteria may instead arrive "late" to the tumor microenvironment and feed on tumors that have alread begun to develop (Flynn 2016). Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found both on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis.

@@

Microbiome-based diagnostics are increasingly being explored as non-invasive tools to survey for the development of colon cancer. Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors. Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit (Baxter et al 2016) and it performed much better than FIT alone. In this study, all of our subjects were tested using FIT and all but two were found to be below the limit of detection (data not shown). Further work from our lab has shown that microbiome profiling of the FIT cartridge contents sufficiently represented the stool community (cite Niel fit paper). Our work already showed that the stool most accurately matched the samples from the left lumen. So we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities. But further studies would need to be done. Hard to get a complete set of matched samples from CRC patients though due to discomfort and insurance issues. One caveat of the previous study is that there was not sufficient information to test if a classification model could differentiate between proximal and distal CRC tumors based on exit stool sample alone, but we would hypothesize that would not be effective.

offer another model/schematic?

-what does this answer tell us about the broader field?
This answer told us that there are specific differences of bacteria in each side and that we can use this information to form hypotheses about how specific microbial/ecological variations from the host can affect oncogenesis. It tells us too that sample collection methods are important, and what we have found in unprepped bowels do/do not match published findings.

By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypothesies about how specific host-microbe interactions can affect oncogenesis of proximal and distal CRC tumors. To this point, 16S community profiling studies are a good start but not enough information to probe these questions. Our sample set of matched proximal, distal, lumenal and mucosal samples from colons that have not undergone bowel preparation presents a unique opportunity to explore further Future studies examining metagenomic, metabolomic and host interactions at each site will be useful in further characterizing the host-microbe interactions in the development of proximal and distal colorectal cancer.

@@

### Methods
Study participants/IRB

Sample collection

Sample processing/DNA extraction

16S rRNA sequencing

Statistical analysis/Random Forest

>>>>

#### Bacterial OTUs associated with cancer are found in healthy individuals 
Some of the healthy patients have Fusobacterium OTUs. The most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subspecies animalis_. It is the only OTU to align with a nucleatum strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. ((cite joe and niels work somewhere in here, maybe in the discussion?)). The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample. ((Table or something?)). OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides. Samples that were positive for _P. asacharolytica_ ranged in abundances from 0.01% - 16%. Finally, another commonly cancer-associated species, _Parvimonas micra_ was identified to be most similar to OTU248. Only 4 individuals had this bacteria present in their colons, three in the left mucosa and one in the right mucosa. Thus, potentially oncogenic OTUs can be found in our sample set of 20 healthy individuals.

@@

We detected _F. nucleatum_ and _P. asacharolytica_ in 7 and 5 of our subjects, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans (cite Niel) and have oncogenic properties in cell culture and in mice (cite). Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the left mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous left mucosa (cite something, Albenberg, 2014). In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on right-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum (cite Mima 2016). Of the (8) (40%) individuals positive for _F. nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumen and left mucosa. That said, in a previous study of 1,102 colorectal cancer cases, only 138 (11%) of them contained detectable _F. nucleatum_ at all (Mima 2016). In a separate study of patients with inflammatory bowel disease (IBD), 50% of 22 IBD patients and 17.6% of 34 people without disease contained _F. nucleatum_ in mucosal biopsies (Strauss 2011). These studies have the added caveat of being sampled from prepared or cleaned colons. Thus, the presence of _F. nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers. Because of the spatial distribution of the _F. nucleatum_ in our sample set, we cannot develop a model for the role of _F. nucleatum_ in the healthy colon. Indeed, rather than the _F. nucleatum_ initiating tumorigenesis, these bacteria may instead arrive "late" to the tumor microenvironment and feed on tumors that have alread begun to develop (Flynn 2016). Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found both on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis.

@@

Microbiome-based diagnostics are increasingly being explored as non-invasive tools to survey for the development of colon cancer. Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors. Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit (Baxter et al 2016) and it performed much better than FIT alone. Further work from our lab has shown that microbiome profiling of the FIT cartridge contents sufficiently represented the stool community (cite Niel fit paper). Our work already showed that the stool most accurately matched the samples from the left lumen. So we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities. But further studies would need to be done. Hard to get a complete set of matched samples from CRC patients though due to discomfort and insurance issues. One caveat of the previous study is that there was not sufficient information to test if a classification model could differentiate between proximal and distal CRC tumors based on exit stool sample alone, but we would hypothesize that would not be effective.

offer another model/schematic? what does this tell us about the broader field? do i need to tie in with the host any more here? (for instance, how microbes might interact with mutated genes/pathways). model, specific microbes on each side, change in mutation or metabolism affects change in community or precedes oncogenesis in a specific way?

By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypothesies about how specific host-microbe interactions can affect oncogenesis of proximal and distal CRC tumors. To this point, 16S community profiling studies do not provide enough information to probe these questions. Our sample set of matched proximal, distal, lumenal and mucosal samples from colons that have not undergone bowel preparation presents a unique opportunity to explore further questions about the microbiome along the gut axis. Specifically, examining metagenomic, metabolomic and host interactions at each site will be useful in further characterizing the host-microbe interactions in the development of proximal and distal colorectal cancer.

@@

### Methods
#### Human subjects
The procedures in this study and consent were approved by the Institutional Review Board at the University of Michigan Health System with protocol number XXXX. Subjects were recruited using the online recruitment platform and were pre-screened prior to enrollment in the study. Exclusion criteria included: use of asprin or NSAIDs within 7 days, use of antibiotics within 3 months, current use of anticoagulants, known allergies to Fentanyl or Benadryl, prior history of colon disease, diabetes, abdominal surgery, respiratory, liver, kidney or brain impairments, undergoing current chemotherapy or radiation treatment and subjects that were pregnant or trying to conceive. 20 subjects that met the criteria were selected and provided signed informed consent prior to the procedure. (Table of participant characteristics? gender, race split?)

#### Sample collection
At a baseline visit, subjects were consented and given a home collection stool kit (Source of kit supplies). At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and two swabs of a Fecal Immunohistochemical Test cartridge (Polymedco) and ship the samples to the University. Notably, subjects did not undergo any bowel preparation method prior to sampling. On procedure day, subjects reported to the Michigan Clinical Research Unit at the University of Michigan Health System. Patients were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate. A flexible sigmoidoscope was first inserted about 25 cm into the colon and endoscopy brush used to collect lumenal/stool contents. Two lumenal samples were collected and the contents immediately deposited into RNAlater (source) and flash-frozen in liquid nitrogen. The brushes were withdrawn and biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter. Three (double check this) mucosal biopsies were collected and flash-frozen in RNAlater. These samples comprised the distal or left colon samples. The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the ascending colon. Samples were then collected as in the distal colon and the colonoscope withdrawn. All samples were stored at -80 C until study completion.

#### Sample processing and sequencing

#### Statistical analysis/Random Forest
